---
comments: false
---

## Defining the data

The key to simulating data in `simstudy` is the creation of series of data defintion tables that look like this:

```{r,  echo=FALSE}
def <- defData(varname = "nr", dist = "nonrandom", formula=7, id = "idnum")
def <- defData(def,varname="x1", dist="uniform", formula="10;20")
def <- defData(def,varname="y1", formula="nr + x1 * 2", variance=8)
def <- defData(def,varname="y2", dist="poisson", formula="nr - 0.2 * x1",link="log")
def <- defData(def,varname="xCat",formula = "0.3;0.2;0.5", dist="categorical")
def <- defData(def,varname="g1", dist="gamma", formula = "5+xCat", variance = 1, link = "log")
def <- defData(def, varname = "a1", dist = "binary" , formula="-3 + xCat", link="logit")

knitr::kable(def)
```

These *definition* tables can be generated two ways. One option is to to use any external editor that allows the creation of `csv` files, which can be read in with a call to `defRead`. An  alternative is to make repeated calls to the function `defData`. Here, we illustrate the R code that builds this definition table internally:

```{r, tidy = TRUE}
def <- defData(varname = "nr", dist = "nonrandom", formula=7, id = "idnum")
def <- defData(def,varname="x1",dist="uniform",formula="10;20")
def <- defData(def,varname="y1",formula="nr + x1 * 2",variance=8)
def <- defData(def,varname="y2",dist="poisson",formula="nr - 0.2 * x1",link="log")
def <- defData(def,varname="xCat",formula = "0.3;0.2;0.5",dist="categorical")
def <- defData(def,varname="g1", dist="gamma", formula = "5+xCat", variance = 1, link = "log")
def <- defData(def, varname = "a1", dist = "binary" , formula="-3 + xCat", link="logit")
```

The first call to `defData` without specifying a definition name (in this example the definition name is *def*) creates a **new** data.table with a single row. An additional row is added to the table `def` each time the function `defData` is called. Each of these calls is the definition of a new field in the data set that will be generated. In this example, the first data field is named 'nr', defined as a constant with a value to be 7. In each call to `defData` the user defines a variable name, a distribution (the default is 'normal'), a mean formula (if applicable), a variance parameter (if applicable), and a link function for the mean (defaults to 'identity').

The possible distributions include **normal**, **gamma**, **poisson**, **zero-truncated poisson**, **binary**, **uniform**, **categorical**, and **deterministic/non-random**. For all of these distributions, key parameters defining the distribution are entered in the `formula`, `variance`, and `link` fields. 

In the case of the **normal** and **gamma** distributions, the formula specifies the mean. The formula can be a scalar value (number) or a string that represents a function of previously defined variables in the data set definition (or, as we will see later, in a previously generated data set). In the example, the mean of `y1`, a normally distributed value, is declared as a linear function of `nr` and `x1`, and the mean of `g1` is a function of the category defined by `xCat`. The `variance` field is defined only for normal and gamma random variables, and can only be defined as a scalar value. In the case of gamma random variables, the value entered in variance field is really a dispersion value $d$, where the actual variance will be $d \times mean^2$.

In the case of the **poisson**, **zero-truncated poisson**, and **binary** distributions, the formula also specifies the mean. The variance is not a valid parameter in these cases, but the `link` field is. The default link is 'identity' but a 'log' link is available for the poisson distributions and a "logit" link is available for the binary outcomes. In this example, `y2` is defined as poisson random variable with a mean that is function of `nr` and `x1` on the log scale. For binary variables, which take a value of 0 or 1, the formula represents probability (with the 'identity' link) or log odds (with the 'logit' link) of the variable having a value of 1. In the example, `a1` has been defined as a binary random variable with a log odds that is a function of `xCat`.

Variables defined with a **uniform**, **categorical**, or **deterministic/non-random** distribution are specified using the formula only. The `variance` and `link` fields are not used in these cases.

For a uniformly distributed variable, The formula is a string with the format "a;b", where *a* and *b* are scalars or functions of previously defined variables. The uniform distribution has two parameters - the minimum and the maximum. In this case, *a* represents the minimum and *b* represents the maximum.

For a categorical variable with \(k\) categories, the formula is a string of  probabilities that sum to 1: "\(p_1 ; p_2 ; ... ; p_k\)". \(p_1\) is the probability of the random variable falling category 1, \(p_2\) is the probablity of category 2, etc. The probabilities can be specified as functions of other variables previously defined. In the example, `xCat` has three possibilities with probabilites 0.3, 0.2, and 0.5, respectively.

Non-random variables are defined by the formula. Since these variables are deterministic, variance is not relevant. They can be functions of previously defined variables or a scalar, as we see in the sample for variable defined as `nr`.

## Generating the data

After the data set definitions have been created, a new data set with \(n\) observations can be created with a call to function **`genData`**. In this example, 1,000 observations are generated using the data set defitions in **`def`**, and then stored in the object **`dt`**:

```{r, tidy = TRUE}
dt <- genData(1000, def)
dt
```

New data can be added to an existing data set with a call to function **`addColumns`**. The new data definitions are created with a call to **`defData`** and then included as an argument in the call to **`addColumns`**:

```{r, tidy = TRUE}
addef <- defDataAdd(varname = "zExtra", dist = "normal", formula = '3 + y1', 
                 variance = 2)

dt <- addColumns(addef, dt)
dt
```
