---
title: 'simstudy: another way to generate data from a non-standard density'
author: Package Build
date: '2024-06-04'
slug: []
categories: []
tags:
  - R
  - simstudy
type: ''
subtitle: ''
image: ''
---

One of my goals for the `simstudy` package is to make it as easy as possible to generate data from a wide range of data distributions. The recent [update](https://www.rdatagen.net/post/2024-05-21-simstudy-customized-distributions/){target="_blank"} created the possibility of generating data from a customized distribution specified in a user-defined function. Last week, I added two functions, `genDataDist` and `addDataDist`, that allow data generation from an empirical distribution defined by a vector of integers. (See [here](https://kgoldfeld.github.io/simstudy/dev/index.html){target="_blank"} for how to download latest development version.) This post provides a simple illustration of the new functionality.

Here are the libraries needed, in case you want to follow along:

```{r}
library(simstudy)
library(data.table)
library(ggplot2)

set.seed(1234)
```

The target density is simply defined by specifying a vector that is intended to loosely represent a data distribution. We start by specifying the vector (which can be of any length):

```{r}
base_data <- 
  c(1, 2, 2, 2, 2, 2, 2, 3, 3, 4, 4, 5, 6, 6, 7, 8, 9, 9, 9, 10, 10, 10, 10, 10)
```

We can look at the density to make sure this is the distribution we are interested in drawing our data from:

```{r, fig.width = 8, fig.height = 4}
emp_density <- density(base_data, n = 10000)
den_curve <- data.table(x = emp_density$x, y = emp_density$y)

ggplot(data = den_curve, aes(x = x, y = y)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(name = "density\n", limits = c(0, 0.11), 
     breaks = seq(0, .10, by = .02)) +
  scale_color_manual(values = colors) +
  theme(panel.grid = element_blank())
```

Actually drawing samples from this distribution is a simple call to `genDataDensity`. The key argument is the data distribution as as represented by the vector of integers:  

```{r, fig.width = 8, fig.height = 4, warning=FALSE}
dx <- genDataDensity(10000, dataDist = base_data, varname = "x1")
```

Here's a look at the sampled data and their relationship to the target density:

```{r, warning=FALSE, fig.height = 4, fig.width = 8}
ggplot(data = dx, aes(x=x1)) +
  geom_histogram(aes(y = after_stat(count / sum(count))), 
    binwidth = 1, fill = "grey", color = "black", alpha = .2) +
  geom_line(data = den_curve, aes(x = x, y = y), 
    color = "black", linewidth = 2) +
  scale_y_continuous(name = "density\n", limits = c(0, 0.11), 
     breaks = seq(0, .10, by = .02)) +
  scale_x_continuous(limits = c(-6, 15), breaks = seq(-5, 10, by = 5)) +
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 10))
```

Just to show that this was not a fluke, here are three additional target distributions, specified with three different vectors:

```{r}
base_data <- list(
  c(1, 1, 1, 1, 1, 2, 2, 2, 2, 3, 3, 3, 3, 4, 4, 5, 5, 6, 6, 7,   7,  8,  9, 10),
  c(1, 2, 2, 3, 3, 4, 4, 4, 4, 4, 4, 5, 6, 6, 7, 7, 7, 8, 9, 10, 10, 10, 10, 10),
  c(1, 2, 3, 4, 4, 5, 5, 6, 6, 7, 7, 8, 8, 8, 8, 9, 9, 9, 9, 10, 10, 10, 10, 10)
)
```

```{r, fig.width = 8, fig.height = 4, echo=FALSE}
den_dt <- function(x) {
  de <- density(x, n = 10000)
  de <- data.table(x = de$x, y = de$y)
} 

dens <- rbindlist(lapply(base_data, function(x) den_dt(x)), idcol = TRUE)

colors <- c("#83d7ea", "#9683ea", "#ea9683")

ggplot(data = dens, aes(x = x, y = y, group = .id)) +
  geom_line(aes(color = factor(.id)), linewidth = 1) +
  scale_y_continuous(name = "density\n", limits = c(0, 0.15), 
    breaks = seq(0, .14, by = .02)) +
  scale_color_manual(values = colors) +
  theme(panel.grid = element_blank(),
        legend.position = "none")
```

We can generate data from each of the distributions and then confirm that each one adequately fits its target distribution:

```{r}
dx1 <- genDataDensity(10000, dataDist = base_data[[1]], varname = "x1")
dx2 <- genDataDensity(10000, dataDist = base_data[[2]], varname = "x1")
dx3 <- genDataDensity(10000, dataDist = base_data[[3]], varname = "x1")
```

```{r, fig.width = 10, fig.height = 2.75, warning=FALSE, echo = FALSE}
dx <- rbindlist(list(dx1, dx2, dx3), idcol = TRUE)

cuts <- seq(dx[,floor(min(x1))], dx[,ceiling(max(x1))], by = 1)
dcuts <- data.table(bin = 1:length(cuts), binlab = cuts)

dx[, bin := cut(x1, breaks = cuts, labels = FALSE)]
dx <- merge(dx, dcuts, by = "bin")

dp <- dx[, .N, keyby = .(.id, binlab)]
dp[, p := N/sum(N), keyby = .id]

ggplot(data = dp, aes(x = binlab, y = p)) +
  geom_bar(aes(fill = factor(.id), color = factor(.id)), stat = "identity", alpha = .4) +
  geom_line(data = dens, aes(x = x, y = y, color = factor(.id)), 
            linewidth = 2) +
  xlab("\nx1") +
  ylab("density\n") +
  scale_fill_manual(values = colors) +
  scale_color_manual(values = colors) +
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 10),
        legend.position = "none") + 
  facet_grid(~ .id)  
```


### Addendum: code to generate multiple distribution plot

Here is a little more detail in case someone might find it useful to have the code that generates the "facet" plot. In the plot with the single distribution, I specified the histogram with this command:

```{r, eval=FALSE}
geom_histogram(aes(y = after_stat(count / sum(count))), ...)
```               

When I tried to apply this to the "facet" plot, the denominator of that plot (`sum(count)`) was not calculated for each subgroup (i.e., dataset), but was the total across all datasets. As a result, the dataset-specific proportions were underestimated; we can see that here:

```{r, warning=FALSE, fig.height = 2.75, fig.width = 10}

dx <- rbindlist(list(dx1, dx2, dx3), idcol = TRUE)

ggplot(data = dx, aes(x=x1)) +
  geom_histogram(
    aes(y = after_stat(count / sum(count)), fill = factor(.id), color = factor(.id)),
    binwidth = 1, alpha = .2) +
  geom_line(data = dens, aes(x = x, y = y, color = factor(.id)), linewidth = 2) +
    xlab("\nx1") +
  ylab("density\n") +
  scale_fill_manual(values = colors) +
  scale_color_manual(values = colors) +
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 10),
        legend.position = "none") + 
  facet_grid(~ .id)  
```

I looked around for a way to address this, but couldn't find anything that obviously addressed this shortcoming (though I am convinced it must be possible, and I just couldn't locate the solution). I considered using `ggarrange`or something similar, but was not satisfied with the results. Instead, it turned out to be faster just to calculate the proportions myself. This is the process I used:

First, I created a dataset with the bins (using a bin size of 1):

```{r}
cuts <- seq(dx[,floor(min(x1))], dx[,ceiling(max(x1))], by = 1)
dcuts <- data.table(bin = 1:length(cuts), binlab = cuts)

dcuts
```

Then, I allocated each observation to a bin using the `cut` function:

```{r}
dx[, bin := cut(x1, breaks = cuts, labels = FALSE)]
dx <- merge(dx, dcuts, by = "bin")

dx
```

Finally, I calculated the distribution-specific proportions (showing only the second distribution):

```{r}
dp <- dx[, .N, keyby = .(.id, binlab)]
dp[, p := N/sum(N), keyby = .id]

dp[.id == 2]
```

And now the facet plot will work just fine. Here is the code and the plot (again).

```{r, fig.width = 10, fig.height = 2.75, warning=FALSE}
ggplot(data = dp, aes(x = binlab, y = p)) +
  geom_bar(aes(fill = factor(.id), color = factor(.id)), stat = "identity", alpha = .4) +
  geom_line(data = dens, aes(x = x, y = y, color = factor(.id)), 
            linewidth = 2) +
  xlab("\nx1") +
  ylab("density\n") +
  scale_fill_manual(values = colors) +
  scale_color_manual(values = colors) +
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 10),
        legend.position = "none") + 
  facet_grid(~ .id)  
```

### Addendum follow-up

Well, that was quick. Andrea provided code on *Disqus* - which for some reason is no longer publishing on my site, and if anyone has thoughts about that issue, feel free to contact me :) - that does exactly what I was trying to do without any pre-plotting data transformations. The trick is to use the *density* stat available in `geom_histogram`, This actually looks better, because it lines up more precisely with the density curve.

```{r, fig.width = 10, fig.height = 2.75, warning=FALSE}
ggplot(data = dx, aes(x=x1)) +
  geom_histogram(
    aes(y = after_stat(density), fill = factor(.id), color = factor(.id)),
    binwidth = 1, alpha = .2) +
  geom_line(data = dens, aes(x = x, y = y, color = factor(.id)), linewidth = 2) +
  xlab("\nx1") +
  ylab("density\n") +
  scale_fill_manual(values = colors) +
  scale_color_manual(values = colors) +
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 10),
        legend.position = "none") +
  facet_grid(~ .id)
```
  
