---
title: Adding competing risks in survival data generation
author:
date: '2022-03-15'
slug: []
categories: []
tags:
  - R
  - survival analysis
type: ''
subtitle: ''
image: ''
---

```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(survminer)
library(ggplot2)

cbbPalette <- c("#B84226","#B88F26", "#A5B435", "#1B8446",
                "#B87326","#B8A526", "#6CA723", "#1C5974") 

ggtheme <- function(panelback = "white") {
  
  ggplot2::theme(
    panel.background = ggplot2::element_rect(fill = panelback),
    panel.grid = ggplot2::element_blank(),
    axis.ticks =  ggplot2::element_line(colour = "black"),
    panel.spacing = ggplot2::unit(0.25, "lines"),  # requires package grid
    panel.border = ggplot2::element_rect(fill = NA, colour="gray90"), 
    plot.title = ggplot2::element_text(size = 8,vjust=.5,hjust=0),
    axis.text = ggplot2::element_text(size=8),
    axis.title = ggplot2::element_text(size = 8)
  )  
  
}
```

I am working on an update of `simstudy` that will make generating survival/time-to-event data a bit more flexible. There are two biggish enhancements. The first facilitates generation of competing events, and the second allows for the possibility of generating survival data that has time-dependent hazard ratios. This post focuses on the first enhancement, and a follow up will provide examples of the second. (If you want to try this at home, you will need the development version of `simstudy`, which you can install using **devtools::install_github("kgoldfeld/simstudy")**.)

### Competing risks

In standard survival analysis, we are interested in understanding the distribution of time to a single event, such as time to death or time to symptom-free recovery. However, there may be situations where two there are at least two possible events, and the occurrence of one precludes the occurrence of the other. In this case, we say that the events are competing risks. In the work that motivated this `simstudy` enhancement, we are trying to model the time to opioid overdose death for people being released from jail, but there are at least two other competing risks, death from other causes and re-incarceration. In addition, there is censoring, which from a data generation point of view (though maybe not from an analytic perspective), can also be considered a competing risk.

### Generating competing risk outcomes under 0.4.0

Previously, it was possible to generate competing risk outcomes, but it had to be done "manually" with additional lines of code or data definitions. It was a bit cumbersome:

```{r, message = FALSE}
library(simstudy)
library(data.table)
library(survival)
```

First, we would generate a data set with `genSurv`:

```{r}
d1 <- defData(varname = "x1", formula = .5, dist = "binary")
d1 <- defData(d1, "x2", .5, dist = "binary")

dS <- defSurv(varname = "event_1", formula = "-12 - 0.1*x1 - 0.2*x2", shape = 0.3)
dS <- defSurv(dS, "event_2", "-12 - 0.3*x1 - 0.2*x2", shape = 0.3)
dS <- defSurv(dS, "event_3", "-12 - 0.4*x1 - 0.3*x2", shape = 0.3)
dS <- defSurv(dS, "censor", "-13", shape = 0.3)

set.seed(2140)

dtCov <- genData(3001, d1)
dtSurv <- genSurv(dtCov, dS)

head(dtSurv)
```

And then we would add the observed time variable with a new data definition and call to `addColumns`:

```{r}
f <- "(time==censor)*0 + (time==event_1)*1 + (time==event_2)*2 + (time==event_3)*3"

cdef <- defDataAdd(varname = "time", 
  formula = "pmin(censor, event_1, event_2, event_3)", dist = "nonrandom")
cdef <- defDataAdd(cdef, varname = "event", 
  formula = f, 
  dist = "nonrandom")

dtSurv_final <- addColumns(cdef, dtSurv)
head(dtSurv_final)
```

### Streamlined approach

The function `addCompRisk` can now generate an observed time variable (which will be the first of the four event times). This is done by specifying a *timeName* argument that will represent the observed time value. The event status is indicated in the field set by the *eventName* argument (which defaults to "event"). And, if a variable name is indicated in the *censorName* argument, the censored events automatically have a value of 0. 

```{r}
dtSurv_final <- addCompRisk(dtSurv, 
  events = c("event_1", "event_2", "event_3", "censor"), 
  timeName = "time", censorName = "censor")

head(dtSurv_final)
```

The competing risk data can be plotted using the cumulative incidence functions (rather than the survival curves):

```{r, tidy = TRUE, echo = FALSE, fig.width = 6.5, fig.height = 3.5, warning=FALSE}
fit <- survfit(Surv(time, event, type="mstate") ~ x1 + x2, data=dtSurv_final)
survminer::ggcompetingrisks(fit, ggtheme = ggtheme("grey94"))  + 
  ggplot2::scale_fill_manual(values = cbbPalette)
```

The data generation can be streamlined even further, as there is no need to call `addCompRisk` at all. The same arguments can be provided directly in the call to `genSurv`, so everything is done at once. And there is no need to explicitly name the events, as they are extracted from the survival distribution definitions:

```{r}
dtSurv_final <- genSurv(dtCov, dS, timeName = "time", censorName = "censor")
head(dtSurv_final)
```

If you don't want to generate competing risk data, and you can set *timeName* to `NULL`, or just not specify it:

```{r}
dtSurv_final <- genSurv(dtCov, dS)
head(dtSurv_final)
```

