---
title: 'simstudy: customized distributions'
author: Package Build
date: '2024-05-21'
slug: []
categories: []
tags:
  - R
  - simstudy
type: ''
subtitle: ''
image: ''
draft: true
---

```{r chunkname, echo=-1}
data.table::setDTthreads(2)
```

```{r, echo = FALSE, message = FALSE}
library(simstudy)
library(ggplot2)
library(scales)
library(grid)
library(gridExtra)
library(survival)
library(gee)
library(data.table)
library(ordinal)

odds <- function (p)  p/(1 - p) # TODO temporary remove when added to package
plotcolors <- c("#B84226", "#1B8445", "#1C5974")

cbbPalette <- c("#B84226","#B88F26", "#A5B435", "#1B8446",
                "#B87326","#B8A526", "#6CA723", "#1C5974") 

ggtheme <- function(panelback = "white") {
  
  ggplot2::theme(
    panel.background = element_rect(fill = panelback),
    panel.grid = element_blank(),
    axis.ticks =  element_line(colour = "black"),
    panel.spacing =unit(0.25, "lines"),  # requires package grid
    panel.border = element_rect(fill = NA, colour="gray90"), 
    plot.title = element_text(size = 8,vjust=.5,hjust=0),
    axis.text = element_text(size=8),
    axis.title = element_text(size = 8)
  )  
  
}

```

Over the past few years, a number of folks have asked if `simstudy` accommodates customized distributions. There's been interest in truncated, zero-inflated, or even more standard distributions that have not yet been implemented in `simstudy`. While I've come up with approaches for some of the specific cases, I was never able to develop a general solution that could provide broader flexibility.

That changes with the latest version of `simstudy` available on [CRAN](https://cran.r-project.org/web/packages/simstudy/index.html){target="_blank"}. Looking through the [issue](https://github.com/kgoldfeld/simstudy/issues){target="_blank"} list recently, I had a brief moment of inspiration, so I decided to tackle the problem anew. I am happy to say that custom distributions can now be specified in `defData` and `defDataAdd` by setting the argument *dist* to "custom". While not perfect, the new functionality has the potential to be quite useful. To introduce the new option, I am providing a couple of examples. 

### Specifying the customized distribution

When defining a custom distribution in `defData`, you provide the name of the user-defined function as a string in the *formula* argument. The arguments of this custom function need to be listed in the *variance* argument, separated by commas and formatted as "**arg_1 = val_form_1, arg_2 = val_form_2, $\dots$, arg_K = val_form_K**".

Here, the *arg_k's* represent the names of the arguments passed to the customized function, where $k$ ranges from $1$ to $K$. You can use values or formulas for each *val_form_k*. If formulas are used, ensure that the variables have been previously generated. Double dot notation is available in specifying *value_formula_k*. One important requirement for the custom function is that the parameter list must include an argument"**n = n**", but do not include $n$ in the definition as part of `defData` or `defDataAdd`.

### Example 1

Here is an example where we would like to generate data from a zero-inflated beta distribution (I've implemented something like this in the past using a mixture distribution, which is a fine way to go). I've created a user-defined function `zeroBeta` that takes on shape parameters $a$ and $b$ for the beta distribution, as well as $p_0$, the proportion of the sample that is zero. Note that the function also takes an argument $n$ that will not to be be specified in the data definition; $n$ will represent the number of observations being generated:

```{r}
zeroBeta <- function(n, a, b, p0) {
  betas <- rbeta(n, a, b)
  is.zero <- rbinom(n, 1, p0)
  betas*!(is.zero)
}
```

The data definition specifies a new variable $zb$ with $a$ and $b$ set to 0.75, and $p_0 = 0.02$:

```{r}
def <- defData(
  varname = "zb", 
  formula = "zeroBeta", 
  variance = "a = 0.75, b = 0.75, p0 = 0.02", 
  dist = "custom"
)
```

The data are generated with a call to `genData` as is typically done in `simstudy`:

```{r}
set.seed(1234)
dd <- genData(100000, def)
```

```{r, echo = FALSE}
dd
```

A plot of the data highlights an over-representation of zeroes:

```{r, fig.width = 6, fig.height = 3, echo = FALSE}
ggplot(data = dd, aes(x = zb)) +
  geom_histogram(binwidth = 0.01, boundary = 0, fill = "grey60") +
  theme(panel.grid = element_blank()) 
```

### Example 2

In this second example, I am generating sets of truncated Gaussian distributions with means ranging from $-1$ to $1$. (I wrote about [this](https://www.rdatagen.net/post/generating-data-from-a-truncated-distribution/){target="_blank"} a while ago - the approach implemented here is an alternative way to generate these data.) `rnormt` is a customized (user-defined) function that generates the truncated Gaussian data, and requires four arguments (the left truncation value, the right truncation value, the distribution average and the standard deviation):

```{r}
rnormt <- function(n, min, max, mu, s) {
  
  F.a <- pnorm(min, mean = mu, sd = s)
  F.b <- pnorm(max, mean = mu, sd = s)
  
  u <- runif(n, min = F.a, max = F.b)
  qnorm(u, mean = mu, sd = s)
  
}
```


In this example, truncation limits differ based on group membership. Initially, three groups are created, followed by the generation of truncated values. For Group 1, truncation is defined by the range of $-1$ to $1$, for Group 2, it's $-2$ to $2$ and for Group 3, it's $-3$ to $3$. We'll generate three data sets, each with a distinct mean denoted by M, using the double-dot notation to implement these different means.

```{r}
def <-
  defData(
    varname = "limit", 
    formula = "1/4;1/2;1/4",
    dist = "categorical"
  ) |>
  defData(
    varname = "tn", 
    formula = "rnormt", 
    variance = "min = -limit, max = limit, mu = ..M, s = 1.5",
    dist = "custom"
  )
```

The data generation requires three calls to `genData`, which I implement with `lapply`:

```{r}
mus <- c(-1, 0, 1)
dd <-lapply(mus, function(M) genData(100000, def))
```

The output is a list of three data sets; here are the first six observations from each of the three data sets:

```{r, echo=FALSE}
lapply(dd, function(D) head(D))
```

A plot highlights the group differences for each of the three data sets:

```{r, fig.width = 8, fig.height = 6, echo = FALSE}
pfunc <- function(dx, i) {
  ggplot(data = dx, aes(x = tn)) +
    geom_histogram(aes(fill = factor(limit)), binwidth = 0.05, boundary = 0, alpha = .8) +
    facet_grid( ~ limit) +
    ggtheme() +
    theme(legend.position = "none") +
    scale_fill_manual(values = plotcolors) +
    scale_x_continuous(breaks = seq(-3, 3, by =1)) +
    scale_y_continuous(limits = c(0, 1000)) +
    ggtitle(paste("mu =", mus[i]))
}

plist <- lapply(seq_along(dd), function(a) pfunc(dd[[a]], a))
grid.arrange(grobs = plist, nrow = 3)
```
