---
title: Planning for a 3-arm cluster randomized trial with a nested intervention and a time-to-event outcome
author: Package Build
date: '2025-05-13'
slug: []
categories: []
tags:
  - R
  - Cluster randomized trials
type: ''
subtitle: ''
image: ''
draft: TRUE
---

A researcher recently approached me for advice on a cluster-randomized trial he was developing. He was interested in testing the effectiveness of two interventions and wondered whether a 2×2 factorial design might be the best approach.

As we discussed the interventions (I'll call them $A$ and $B$), it became clear that $A$ was the primary focus. Intervention $B$ might enhance the effectiveness of $A$, but on its own, $B$ was not expected to have much impact. (It's also possible that $A$ alone doesn't work, but once $B$ is in place, the combination may reap benefits.) Given this, it didn’t seem worthwhile to randomize clinics or providers to receive B alone. We agreed that a three-arm cluster-randomized trial---with (1) control, (2) $A$ alone, and (3) $A + B$---would be a more efficient and relevant design.

A while ago, I [wrote about](https://www.rdatagen.net/post/2023-12-19-a-three-arm-trial-using-two-step-randomization/){target="_blank"} a proposal to conduct a three-arm trial using a two-step randomization scheme. That design assumes that outcomes in the enhanced arm ($A + B$) are uncorrelated with those in the standalone arm  $A$ within the same cluster. For this project, that assumption didn’t seem plausible, so I recommended sticking with a standard cluster-level randomization.

The study has three goals:

* Assess the effectiveness of $A$ versus control
* Compare $A + B$ versus $A$ alone
* If $A$ alone is ineffective, compare $A + B$ versus control

In other words, we want to make three pairwise comparisons. Initially, we were concerned about needing to adjust our tests for multiple comparisons. However, we used a [gatekeeping](https://onlinelibrary.wiley.com/doi/full/10.1002/sim.8463){target="_blank"} strategy that maintains the overall Type I error rate at 5% while allowing each test to be performed at $\alpha = 0.05$.

This post describes how I set up simulations to evaluate sample size requirements for the proposed trial. The primary outcome is a time-to-event measure: the time from an index physician visit to a follow-up visit, which the intervention aims to shorten. I first generated survival data based on estimates from the literature, then simulated the study design under various sample size assumptions. For each scenario, I generated multiple datasets and applied the gatekeeping hypothesis tests to estimate statistical power.

### Preliminaries

Before getting started, here are the `R` packages used in this post. In addition, I've set a randomization seed to that if you attempt to replicate the approach taken here, our results should align.

```{r, message=FALSE}
library(simstudy)
library(data.table)
library(survival)
library(coxme)
library(broom)

set.seed(8271)
```

### Generating time-to-event data

#### Get parameters that define survival curve

```{r plot_parameters, fig.height = 4, fig.width = 7}
mq <- c(1.44, 4.68, 8.69, 15.32)

points <- list(c(mq[1], 0.80), c(mq[2], 0.60), c(mq[3], .40), c(mq[4], .2))
r <- survGetParams(points)

survParamPlot(f = r[1], shape = r[2], points, limits = c(0,50))
```

### Generating data for a simpler two-arm RCT

#### Visualize curve and assess its properties

```{r}
def <- defData(varname = "A", formula = "1;1", dist = "trtAssign")

defS <- 
  defSurv(varname = "time", formula = "..int + A * ..eff", shape = "..shape") |>
  defSurv(varname = "censor", formula = -40, scale = 0.5, shape = .10)
```


```{r, warning=FALSE}
int <- r[1]
shape <- r[2]

eff <- log(2)

dd <- genData(100000, def)
dd <- genSurv(dd, defS, timeName = "time", censorName = "censor")

dd[A==0, quantile(time, probs = c(0.20, 0.50, 0.80))]
```


```{r plot_ideal_curves, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=4}
library(survminer)
library(ggplot2)

dd_surv <- survfit(Surv(time, event) ~ A, data = dd)
dd_surv_tidy <- data.table(surv_summary(dd_surv, data = dd))

ggplot(dd_surv_tidy, aes(time, surv, color = factor(A))) +
  geom_step() +  # Kaplan-Meier curves
  labs(x = "Time to event", y = "Probability of no event") +
  scale_color_manual(values = c("#ce9995", "#95aece")) +
  theme(legend.position = "none",
        panel.grid = element_blank()) +
  xlim(0, 20)
```

#### Model fit

```{r}
fit <- coxph(Surv(time, event) ~  factor(A), data = dd)
tidy(fit, exponentiate = TRUE)
```

#### Relationship of HR to median time to event

```{r}
getmedian <- function(eff = 0) {
  
  dd <- genData(100000, def)
  dd <- genSurv(dd, defS, timeName = "time", censorName = "censor")
  dd[A == 1, round(median(time), 1)]
  
}

sapply(log(seq(1, 2, by = .1)), function(x) getmedian(x))
```

### Simulating three-arm study data

#### Data definititions

```{r}
defC <- defData(varname = "b", formula = 0, variance = "..v_clinic")

defP <- 
  defDataAdd(varname = "g", formula = 0, variance = "..v_prov") |>
  defDataAdd(varname = "A", formula = "1;1;1", variance = "clinic", dist = "trtAssign")

defS <- defSurv(
  varname = "eventTime", 
  formula = "..int + b + g + (A==2)*..eff_A + (A==3)*..eff_AB", 
  shape = "..shape") 
```

#### Parameters for data generation

```{r}
nC <- 16               # number of clinics (clusters)
nP <- 6                # number of providers per clinic
nI <- 48               # number of patients per provider
v_clinic <- 0.10       # variation across clinics
v_prov <- 0.25         # variation across providers
eff_A <- log(c(1.4))   # log HR of intervention A (compared to control)
eff_AB <- log(c(1.6))  # log HR of combined A+B (compared to control)
```

#### Data generation

```{r}
ds <- genData(nC, defC, id = "clinic")
dp <- genCluster(ds, "clinic", nP, "provider")
dp <- addColumns(defP, dp)
dd <- genCluster(dp, "provider", nI, "id")
dd <- genSurv(dd, defS)
dd <- trtAssign(dd, nTrt = 12, strata = "provider", grpName = "month")
  
dd[, event := as.integer(eventTime <= 18 - month)]
dd[, obsTime := pmin(eventTime, 18 - month)]
```

```{r plot_provider_km, echo=FALSE}
dq <- copy(dd)
dq$clinic <- factor(dq$clinic)
dq$provider <- factor(dq$provider)
dd_surv <- survfit(Surv(obsTime, event) ~ clinic + provider + A, data = dq)
dd_surv_tidy <- data.table(surv_summary(dd_surv, data = dq))

ggplot(dd_surv_tidy, aes(time, surv, color = factor(A))) +
  geom_step(aes(group=provider)) +  # Kaplan-Meier curves
  facet_wrap(~ clinic, scales = "free_y") +
  labs(x = "Time to event", y = "Probability of no event") +
  scale_color_manual(values = c("#cbce95", "#ce9995", "#95aece")) +
  theme(legend.position = "none",
        panel.grid = element_blank()) +
  xlim(0, 17)
```

```{r}
me.fit <- coxme(
  Surv(eventTime, event) ~ factor(A) + (1|provider) + (1|clinic), 
  data = dd
)

summary(me.fit)
```

```{r plot_provider, echo=FALSE, fig.width=7, fig.height=4}
load("data/power.rda")

ggplot(data = pdata_provider, aes(x = eff_A, y = power, group = diff)) +
  geom_hline(yintercept = 0.8, color = "white") +
  geom_line(color = "grey75") +
  geom_point(aes(color = factor(diff))) +
  scale_color_manual(
    values = c("#ff4500", "#00baff"), 
    name = "HR difference for\ncombined intervention") +
  xlab("Hazard ratio for provider intervention alone") +
  guides(color = guide_legend(reverse = TRUE, title.position = "left")) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.2)) +
  ggtitle("Estimated power: randomization by provider") +
  theme(panel.grid = element_blank(),
        plot.title = element_text(size = 10, face = "bold"),
        legend.title = element_text(size = 8, family = "Verdana"),
        legend.position = "inside",
        legend.position.inside = c(.75, .25),
        legend.box.just = "center"
  )
```

```{r plot_clinic_1, echo=FALSE, fig.width=7, fig.height=4}
ggplot(data = pdata_clinic_1, aes(x = eff_A, y = power, group = diff)) +
  geom_hline(yintercept = 0.8, color = "white") +
  geom_line(color = "grey75") +
  geom_point(aes(color = factor(diff))) +
  scale_color_manual(
    values = c("#ff4500", "#00baff"), 
    name = "HR difference for\ncombined intervention") +
  xlab("Hazard ratio for provider intervention alone") +
  guides(color = guide_legend(reverse = TRUE, title.position = "left")) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.2)) +
  ggtitle("Estimated power: randomization by clinic") +
  theme(panel.grid = element_blank(),
        plot.title = element_text(size = 10, face = "bold"),
        legend.title = element_text(size = 8, family = "Verdana"),
        legend.position = "inside",
        legend.position.inside = c(.75, .25),
        legend.box.just = "center"
  )
```

```{r plot_clinic_2, echo=FALSE, fig.width=7, fig.height=4}
ggplot(data = pdata_clinic_2, aes(x = nC, y = power)) +
  geom_hline(yintercept = 0.8, color = "white") +
  geom_line(color = "grey75") +
  geom_point() +
  xlab("Number of clinics") +
  guides(color = guide_legend(reverse = TRUE, title.position = "left")) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.2)) +
  ggtitle("Estimated power: randomization by clinic") +
  theme(panel.grid = element_blank(),
        plot.title = element_text(size = 10, face = "bold"))
```


<p><small><font color="darkkhaki">
References:

Proschan, M.A. and Brittain, E.H., 2020. A primer on strong vs weak control of familywise error rate. Statistics in medicine, 39(9), pp.1407-1413.

</font></small></p>
