---
title: 'Flexible correlation generation: revisiting block matrices in simstudy'
author: Package Build
date: '2023-02-28'
slug: []
categories: []
tags:
  - R
  - simstudy
  - correlated data
type: ''
subtitle: ''
image: ''
draft: TRUE
---

[Last time](){target="_blank"}, I described enhancements to the `simstudy` function `genCorMat` that are likely to appear in the next version. Here, I am also describing new functionality for generating correlated data, but the focus is on a situation where there is a *temporal* element to the outcome data generation. This is not entirely new. In the previous development version of `simstudy`, I introduced a function `genBlockMat` (which I wrote [about](https://www.rdatagen.net/post/2022-11-22-generating-cluster-data-with-marginal-correlations/){target="_blank"}). Upon further reflection, I wasn't satisfied with the lack of flexibility and clarity of that function, so I've switched gears slightly and am introducing two simpler, but more flexible functions to achieve the same goals: `blockExchangeMat` and `blockDecayMat`.

I provided the motivation for time-dependent correlation matrices in the earlier [post](https://www.rdatagen.net/post/2022-11-22-generating-cluster-data-with-marginal-correlations/){target="_blank"}, so I won't repeat myself here (but if you do read that, don't pay attention to the function I describe there). 

In the update, I've based the parameterization of the data generation process on correlation matrix structures described by [Li *et al*](https://journals.sagepub.com/doi/pdf/10.1177/0962280220932962){target="_blank"}. They've created a very useful nomenclature for laying out different patterns of correlation that can occur within a cluster over time. This classification is based on the distinction between  *cross-sectional* vs *cohort* samples as well as the *exchangeability* vs. *decay* patterns of correlation. 

### Cross-sectional data

In the case where individuals in clusters are "treated" only once over the course of a study, then the sample is considered *cross-sectional*. (This treatment might be the control condition or the experimental condition.) I am assuming that the outcome is always collected relatively close in time to the treatment - this just simplifies things conceptually. However, it does not necessarily to be the case, as long as the patient is exposed only a single time. 

The key point is that in a cross-sectional design, people in the same cluster who are treated at different time periods will be unique. The correlation between individual outcomes will be driven entirely by the intra-cluster correlation. The structure of that intra-cluster correlation will depend on the assumption we make about how correlation changes over time: the correlation can reflect either *exchangeability* or *decay*.

#### Exchangeable

Under the assumption of exchangeability, there is a within-cluster within-period correlation ($\rho_w$) across all study participants in the same cluster and period. For participants in the same cluster but *different* period, the within-cluster between-period correlation ($\rho_b$) is different (presumably lower) but constant over time.

A conceptual diagram of this exchangeable correlation matrix for a cross-sectional design is shown below for a single cluster, and includes three periods and two individuals per. Each box represents a different time period. So, the correlation represented in the box in the upper left hand corner is the within-cluster within-period correlation for the first period. The bottom left box represents the within-cluster between-period correlation for the individuals in the first and third periods. (Note that we are assuming that individuals in different clusters are not correlated with each other.)

<br>

```{r, echo=FALSE}
library(katex)
matform <- "\\scriptsize{

R = \\left ( \\begin{array}{c|c|c}

\\begin{matrix} 
1 & \\rho_w \\\\
\\rho_w & 1
\\end{matrix} &
  
  \\begin{matrix} 
\\rho_b & \\rho_b \\\\
\\rho_b & \\rho_b
\\end{matrix} &
  
  \\begin{matrix} 
\\rho_b & \\rho_b \\\\
\\rho_b & \\rho_b
\\end{matrix} \\\\

\\hline

\\begin{matrix} 
\\rho_b & \\rho_b \\\\
\\rho_b & \\rho_b
\\end{matrix} &
  
  \\begin{matrix} 
1 & \\rho_w \\\\
\\rho_w & 1
\\end{matrix} &
  
  \\begin{matrix} 
\\rho_b & \\rho_b \\\\
\\rho_b & \\rho_b
\\end{matrix} \\\\

\\hline

\\begin{matrix} 
\\rho_b & \\rho_b \\\\
\\rho_b & \\rho_b 
\\end{matrix} &
  
  \\begin{matrix} 
\\rho_b & \\rho_b \\\\
\\rho_b & \\rho_b 
\\end{matrix} &
  
\\begin{matrix} 
1 & \\rho_w \\\\
\\rho_w & 1 
\\end{matrix}

\\end{array} \\right ) }"

katex_html(matform)
```

#### Decay

Under the assumption of decay, the within-cluster within-period correlation ($\rho_w$) is the same as under the exchangeability assumptions. The between-period correlation is now a function of the difference in time when the two individuals were treated. It is $\rho_w * r^{|s-t|}$, where $r$ is a decay parameter between 0 and 1, and $s$ and $t$ are the periods under consideration. For example, in the lower left-hand box, we have the correlation between individuals in the first period ($s=1$) and individuals in the third period ($t=3$), which gives a correlation coefficient of $\rho_w \times r^{|1-3|} = \rho_w \times r^2$. As the difference in periods grows, $r^{|s-t|}$ gets smaller.

<br>

```{r, echo=FALSE}
matform <- "\\scriptsize{

R = \\left ( \\begin{array}{c|c|c}

\\begin{matrix} 
1 & \\rho_w \\\\
\\rho_w & 1
\\end{matrix} &
  
\\begin{matrix} 
\\rho_w r & \\rho_w r \\\\
\\rho_w r & \\rho_w r
\\end{matrix} &
  
\\begin{matrix} 
\\rho_w r^2 & \\rho_w r^2 \\\\
\\rho_w r^2 & \\rho_w r^2
\\end{matrix} \\\\

\\hline

\\begin{matrix} 
\\rho_w r & \\rho_w r \\\\
\\rho_w r & \\rho_w r
\\end{matrix}&
  
  \\begin{matrix} 
1 & \\rho_w \\\\
\\rho_w & 1
\\end{matrix} &
  
  \\begin{matrix} 
\\rho_w r & \\rho_w r \\\\
\\rho_w r & \\rho_w r
\\end{matrix} \\\\

\\hline

\\begin{matrix} 
\\rho_w r^2 & \\rho_w r^2 \\\\
\\rho_w r^2 & \\rho_w r^2
\\end{matrix} &
  
  \\begin{matrix} 
\\rho_w r & \\rho_w r \\\\
\\rho_w r & \\rho_w r
\\end{matrix} &
  
\\begin{matrix} 
1 & \\rho_w \\\\
\\rho_w & 1 
\\end{matrix}

\\end{array} \\right ) }"

katex_html(matform)
```

### Closed cohort

When individuals in clusters are "treated" or exposed in each period of a study, the sample is considered to be a *cohort*. Actually, if every individual is measured in each period, as I've just described, this would be a *closed* cohort, closed in the sense that once the cohort is defined at the beginning of the study, no new participants are added. If we allow participants to start and stop and random points, this would be an *open* cohort design. For the purposes of simulation it is challenging to generate data under an open cohort design with this marginal approach (using correlation matrices), and is much easier to do with random effects (which I did [here](https://www.rdatagen.net/post/simulating-an-open-cohort-stepped-wedge-trial/){target="_blank"}). What I am describing here applies to closed cohorts only.

#### Exchangeable

The key difference between the cross-sectional and cohort design is the within-individual between-period (auto) correlation. Under the exchangeable assumption, the autocorrelation is specified with the correlation coefficient $\rho_a$. The within-period between-individual correlation is still $\rho_w$, and the between-period between-individual correlation is still $\rho_b$. All of these correlations remain constant in the exchangeable framework:

<br>

```{r, echo=FALSE}
matform <- "\\scriptsize{

R = \\left ( \\begin{array}{c|c|c}

\\begin{matrix} 
1 & \\rho_w \\\\
\\rho_w & 1
\\end{matrix} &
  
  \\begin{matrix} 
\\rho_a & \\rho_b \\\\
\\rho_b & \\rho_a
\\end{matrix} &
  
  \\begin{matrix} 
\\rho_a & \\rho_b \\\\
\\rho_b & \\rho_a
\\end{matrix} \\\\

\\hline

\\begin{matrix} 
\\rho_a & \\rho_b \\\\
\\rho_b & \\rho_a
\\end{matrix} &
  
  \\begin{matrix} 
1 & \\rho_w \\\\
\\rho_w & 1
\\end{matrix} &
  
  \\begin{matrix} 
\\rho_a & \\rho_b \\\\
\\rho_b & \\rho_a
\\end{matrix} \\\\

\\hline

\\begin{matrix} 
\\rho_a & \\rho_b \\\\
\\rho_b & \\rho_a 
\\end{matrix} &
  
  \\begin{matrix} 
\\rho_a & \\rho_b \\\\
\\rho_b & \\rho_a 
\\end{matrix} &
  
\\begin{matrix} 
1 & \\rho_w \\\\
\\rho_w & 1 
\\end{matrix}

\\end{array} \\right ) }"

katex_html(matform)
```

#### Decay

The decay structure under an assumption of a closed cohort is the last of the four possible variations. The within-period between-individual correlation $\rho_w$ remains the same, and so does the between-period between-individual correlation $\rho_wr^{|s-t|}$. The between-period within-individual correlation is specified as $r^{|s-t|}$:

<br>

```{r, echo=FALSE}
matform <- "\\scriptsize{

R = \\left ( \\begin{array}{c|c|c}

\\begin{matrix} 
1 & \\rho_w \\\\
\\rho_w & 1
\\end{matrix} &
  
\\begin{matrix} 
r & \\rho_w r \\\\
\\rho_w r & r
\\end{matrix} &
  
\\begin{matrix} 
r^2 & \\rho_w r^2 \\\\
\\rho_w r^2 & r^2
\\end{matrix} \\\\

\\hline

\\begin{matrix} 
r & \\rho_w r \\\\
\\rho_w r & r
\\end{matrix}&
  
  \\begin{matrix} 
1 & \\rho_w \\\\
\\rho_w & 1
\\end{matrix} &
  
  \\begin{matrix} 
r & \\rho_w r \\\\
\\rho_w r & r
\\end{matrix} \\\\

\\hline

\\begin{matrix} 
r^2 & \\rho_w r^2 \\\\
\\rho_w r^2 & r^2
\\end{matrix} &
  
  \\begin{matrix} 
r & \\rho_w r \\\\
\\rho_w r & r
\\end{matrix} &
  
\\begin{matrix} 
1 & \\rho_w \\\\
\\rho_w & 1 
\\end{matrix}

\\end{array} \\right ) }"

katex_html(matform)
```

### Generating block matrices and simulating data

We can put all of this into action with the help of two new functions `blockExchangeMat` and `blockDecayMat`. In the simulations that follow, I will start with a single cluster and three periods, just to keep it simple. In the addendum, I will provide a more elaborate example.


#### Cross-sectional data with exchangeable correlation

```{r}
library(simstudy)
library(data.table)

R_XE <- blockExchangeMat(ninds = 2 , nperiods = 3, rho_w = .5,
  rho_b = .3, pattern = "xsection")

R_XE
```

```{r}
defC <- defData(varname = "lambda", formula = "c(4, 7)", dist = "nonrandom")

set.seed(1234)

dc <- genData(2, defC, id = "site")
dp <- addPeriods(dc, 3, idvars = "site")
dx <- genCluster(dtClust = dp, cLevelVar = "timeID", 
                 numIndsVar = 2, level1ID = "id")

dd <- addCorGen(dx, idvar = "site", corMatrix = R_XE,
  dist = "poisson", param1 = "lambda", cnames = "y")

dd
```

```{r}
replicate <- function(R, dx) {
  reps <- lapply(1:5000, function(x)
    addCorGen(dx, idvar = "site", corMatrix = R,
      dist = "poisson", param1 = "lambda", cnames = "y")
  )

  drep <- data.table::rbindlist(reps, idcol = "rep")
  drep[, seq := 1:.N, keyby = rep]
  dmat <- as.matrix(dcast(drep, rep ~ seq, value.var = "y")[, -1])
  round(cor(dmat), 1) 
}

replicate(R = R_XE, dx = dx)

```

#### Cross-sectional data with correlation decay

```{r}
R_XD <- blockDecayMat(ninds = 2 , nperiods = 3, rho_w = .5,
  r = .8, pattern = "xsection")

R_XD
```

```{r}
replicate(R = R_XD, dx = dx)
```

#### Cohort data with exchangeable correlation

```{r}
R_CE <- blockExchangeMat(ninds = 2 , nperiods = 3, rho_w = .5, 
  rho_b = .3, rho_a = .4, pattern = "cohort")

R_CE
```

```{r}
replicate(R = R_CE, dx = dx)
```

#### Cohort data with correlation decay

```{r}
R_CD <- blockDecayMat(ninds = 2 , nperiods = 3, rho_w = .5, 
  r = .8, pattern = "cohort")

R_CD
```

```{r}
replicate(R = R_CD, dx = dx)
```


<p><small><font color="darkkhaki">
Reference:

Li, Fan, James P. Hughes, Karla Hemming, Monica Taljaard, Edward R. Melnick, and Patrick J. Heagerty. "Mixed-effects models for the design and analysis of stepped wedge cluster randomized trials: an overview." Statistical Methods in Medical Research 30, no. 2 (2021): 612-639.

</font></small></p>

<br>

### Addendum

I want to provide a slightly more elaborate example to show the true flexibility of the new functions. In the case of a cross-sectional design, the number of observations per period for a specific cluster does not need to remain constant (though in the case of data generation under a cohort design it does). We can vary the total number of observations as well as the correlation parameters by cluster.

In this example, there are 10 clusters and three periods. The number of individuals per cluster per period ranges from two to four, and are randomly generated. The decay rate $r$ varies by cluster (generated using the *beta* distribution with shape parameters 6 and 2). The parameter $\rho_w$ is constant across all clusters and is 0.6

```{r}
defC <- defData(varname = "lambda", formula = "sample(5:10, 1)", dist = "nonrandom")
defP <- defDataAdd(varname = "n", formula = "2;4", dist="uniformInt")

dc <- genData(n = 10, dtDefs = defC, id = "site")
dc <- addPeriods(dtName = dc, nPeriods = 3, 
                 idvars = "site", perName = "period")
dc <- addColumns(defP, dc)

dd <- genCluster(dtClust = dc, cLevelVar = "timeID", 
  numIndsVar = "n", level1ID = "id")

```

```{r}
r <- round(rbeta(10, 6, 2), 2)
N <- dd[, .N, keyby = .(site, period)][, N]

R <- blockDecayMat(ninds = N , nperiods = 3, rho_w = 0.6, r = r, nclusters = 10)

reps <- lapply(1:5000, 
  function(x) addCorGen(dd, idvar = "site", corMatrix = R,
    dist = "poisson", param1 = "lambda", cnames = "y")
)

drep <- data.table::rbindlist(reps, idcol = "rep")

empir_corr <- function(cluster) {
  dcrep <- drep[site == cluster, ]
  dcrep[, seq := 1:.N, keyby = rep]
  dmat <- as.matrix(dcast(dcrep, rep ~ seq, value.var = "y")[, -1])
  
  return(round(cor(dmat), 2))
}
```

```{r}
clust <- 1
round(R[[clust]], 2)
empir_corr(cluster = clust)
```

```{r}
clust <- 3
round(R[[clust]], 2)
empir_corr(cluster = clust)
```

```{r}
clust <- 7
round(R[[clust]], 2)
empir_corr(cluster = clust)
```
