---
title: 'simstudy update: two new functions that generate correlated observations from non-normal distributions'
author: ''
date: '2017-07-03'
slug: simstudy-update-two-functions-for-correlation
categories: []
tags:
  - R
subtitle: ''
---
```{r global_options, echo = FALSE, include = FALSE}

options(width = 70)

set.seed(111)

library(simstudy)
library(ggplot2)
library(scales)
library(grid)
library(gridExtra)
library(survival)
library(knitr)
library(gee)
```

In an earlier [post](https://www.rdatagen.net/post/correlated-data-copula/), I described in a fair amount of detail an algorithm to generate correlated binary or Poisson data. I mentioned that I would be updating `simstudy` with functions that would make generating these kind of data relatively painless. Well, I have managed to do that, and the updated package (version 0.1.3) is available for download from [CRAN](https://cran.r-project.org/web/packages/simstudy/index.html). There are now two additional functions to facilitate the generation of correlated data from *binomial*, *poisson*, *gamma*, and *uniform* distributions: `genCorGen` and `addCorGen`. Here's a brief intro to these functions.

### Generate generally correlated data

`genCorGen` is an extension of `genCorData`, which was provided in earlier versions of `simstudy` to generate multivariate normal data. In the first example below, we are generating data from a multivariate Poisson distribution. To do this, we need to specify the mean of the Poisson distribution for each new variable, and then we specify the correlation structure, just as we did with the normal distribution.


```{r firstchunk}
l <- c(8, 10, 12) # lambda for each new variable

dp <- genCorGen(1000, nvars = 3, params1 = l, dist = "poisson", 
                rho = 0.3, corstr = "cs", wide = TRUE)
dp
```

Here is the the estimated correlation (we would expect an estimate close to 0.3):

```{r}
round(cor(as.matrix(dp[, .(V1, V2, V3)])), 2)
```

Similarly, we can generate correlated binary data by specifying the probabilities:

```{r}
db<- genCorGen(1000, nvars = 3, params1 = c(.3, .5, .7), dist = "binary", 
          rho = 0.8, corstr = "cs", wide = TRUE)
db
```

In the case of the binary outcome, the observed correlation  will be lower that what is specified, which in this case was 0.8. I tried to provide some intuition about this in the earlier [post](https://www.rdatagen.net/post/correlated-data-copula/):
```{r}
round(cor(as.matrix(db[, .(V1, V2, V3)])), 2)
```

The gamma distribution requires two parameters - the mean and dispersion. (These are converted into shape and rate parameters more commonly used.)

```{r}
dg <- genCorGen(1000, nvars = 3, params1 = c(3,5,7), params2 = c(1,1,1),
                dist = "gamma", rho = .7, corstr = "cs", 
                wide = TRUE, cnames="a, b, c")
dg

round(cor(as.matrix(dg[, .(a, b, c)])), 2)
```

These data sets can be generated in either *wide* or *long* form. So far, we have generated *wide* form data, where there is one row per unique id. The *long* form, where the correlated data are on different rows, is useful for plotting or fitting models, because  there are repeated measurements for each id:

```{r}
dgl <- genCorGen(1000, nvars = 3, params1 = l, params2 = c(1,1,1), 
                 dist = "gamma", rho = .7, corstr = "cs", wide = FALSE, 
                 cnames="NewCol")
dgl
```

Here is a plot of a subset of the data:

```{r}
ids <- sample(1000,50, replace = FALSE)
ggplot(data=dgl[id %in% ids,], aes(x=factor(period), y=NewCol, group=id)) +
  geom_line(aes(color=factor(id)))+
  theme(legend.position = "none") +
  scale_x_discrete(expand = c(0,0.1))
```

### Generate data based on values from existing data set

`addCorGen` allows us to create correlated data from an existing data set, as one can already do using `addCorData`, but with non-normal data. In the case of `addCorGen`, the parameter(s) used to define the distribution is a field (or fields) in the data set. The correlated data are added to the existing data set. In the example below, we are going to generate three sets (Poisson, binary, and gamma) of correlated data with means that are a function of the variable `xbase`, which varies by id.

First we define the data and generate a data set:

```{r}
def <- defData(varname = "xbase", formula = 5, variance = 0.2, 
               dist = "gamma", id = "cid")
def <- defData(def, varname = "lambda", formula = "0.5 + 0.1 * xbase", 
               dist="nonrandom", link = "log")
def <- defData(def, varname = "p", formula = "-2.0 + 0.3 * xbase", 
               dist="nonrandom", link = "logit")
def <- defData(def, varname = "gammaMu", formula = "0.5 + 0.2 * xbase", 
               dist="nonrandom", link = "log")
def <- defData(def, varname = "gammaDis", formula = 1, 
               dist="nonrandom")

dt <- genData(10000, def)
dt
```

The Poisson distribution has a single parameter, lambda:

```{r}

dtX1 <- addCorGen(dtOld = dt, idvar = "cid", nvars = 3, rho = 0.1, 
                  corstr = "cs", dist = "poisson", param1 = "lambda", 
                  cnames = "a, b, c")

dtX1[, .(cid, xbase, lambda, a, b, c)]
```

The Bernoulli (binary) distribution has a single parameter, p:

```{r}
dtX2 <- addCorGen(dtOld = dt, idvar = "cid", nvars = 4, rho = .4, 
                  corstr = "ar1", dist = "binary", param1 = "p")

dtX2[, .(cid, xbase, p, V1, V2, V3, V4)]
```

And here is the the Gamma distribution, with its two parameters (mean and dispersion):

```{r}
dtX3 <- addCorGen(dtOld = dt, idvar = "cid", nvars = 3, rho = .4, 
                  corstr = "cs", dist = "gamma", 
                  param1 = "gammaMu", param2 = "gammaDis")

dtX3[, .(cid, xbase, gammaMu, gammaDis, 
         V1 = round(V1,2), V2 = round(V2,2), V3 = round(V3,2))]
```

### Long form data

If we have data in *long* form (e.g. longitudinal data), the function will recognize the structure:

```{r}
def <- defData(varname = "xbase", formula = 5, variance = .4, 
               dist = "gamma", id = "cid")
def <- defData(def, "nperiods", formula = 3, 
               dist = "noZeroPoisson")

def2 <- defDataAdd(varname = "lambda", 
                   formula = "0.5 + 0.5 * period + 0.1 * xbase", 
                   dist="nonrandom", link = "log")

dt <- genData(1000, def)

dtLong <- addPeriods(dt, idvars = "cid", nPeriods = 3)
dtLong <- addColumns(def2, dtLong)

dtLong

### Generate the data 

dtX3 <- addCorGen(dtOld = dtLong, idvar = "cid", nvars = 3, 
                  rho = .6, corstr = "cs", dist = "poisson", 
                  param1 = "lambda", cnames = "NewPois")
dtX3
```

We can fit a generalized estimating equation (GEE) model and examine the coefficients and the working correlation matrix. As we would expect, they match closely to the data generating parameters:

```{r}
geefit <- gee(NewPois ~ period + xbase, data = dtX3, id = cid, 
              family = poisson, corstr = "exchangeable")

round(summary(geefit)$working.correlation, 2)

```

In the future, I plan on adding other distributions. Some folks have suggested the negative binomial distribution, which I will do. If you have other suggestions/requests, [let me know](mailto:keith.goldfeld@nyumc.org).