---
comments: false
---

## Generating the treatment/exposure

Treatment assignment can be accomplished through the original data generation process, using `defData` and `genData`. However, the functions `trtAssign` and `trtObserve` provide more options to generate treatment assignment.

### Assigned treatment

Treatment assignment can simulate how treatment is made in a randomized study. Assignment to treatment groups can be (close to) balanced (as would occur in a block randomized trial); this balancing can be done without or without strata. Alternatively, the assignment can be left to chance without blocking; in this case, balance across treatment groups is not guaranteed, particularly with small sample sizes.

First, create the data definition:

```{r, tidy = TRUE, eval = FALSE}
def <- defData(varname = "male", dist = "binary", formula = .5 , id="cid")
def <- defData(def, varname = "over65", dist = "binary", formula = "-1.7 + .8*male", link="logit")
def <- defData(def, varname = "baseDBP", dist = "normal", formula = 70, variance = 40)

dtstudy <- genData(330, def)
```

*Balanced treatment assignment, stratified by gender and age category (not blood pressure)*
```{r, tidy = TRUE, eval = FALSE}
study1 <- trtAssign(dtstudy , n=3, balanced = TRUE, strata = c("male","over65"), grpName = "rxGrp")

study1
```

STUDY 1 TABLE

*Balanced treatment assignment (without stratification)*
```{r, tidy = TRUE, eval = FALSE}
study2 <- trtAssign(dtstudy , n=3, balanced = TRUE, grpName = "rxGrp")
```

STUDY 2 TABLE

*Random (unbalanced) treatment assignment*
```{r, tidy = TRUE, eval = FALSE}
study3 <- trtAssign(dtstudy , n=3, balanced = FALSE, grpName = "rxGrp")
```

STUDY 3 TABLE

*Comparison of three treatment assignment mechanisms*

figbalance

### Observed treatment

If exposure or treatment is observed (rather than randomly assigned), use `trtObserved` to generate groups. There may be any number of possible exposure or treatment groups, and the probability of exposure to a specific level can depend on covariates already in the data set. In this case, there are three exposure groups that vary by gender and age:

```{r, tidy = TRUE, eval = FALSE}
formula1 <- c("-2 + 2*male - .5*over65", "-1 + 2*male + .5*over65")
dtExp <- trtObserve(dtstudy, formulas = formula1, logit.link = TRUE, grpName = "exposure")
```

Here are the exposure distributions by gender and age:

FIG OBS1

Here is a second case of three exposures where the exposure is independent of any covariates. Note that specifying the formula as `c(.35, .45)` is the same as specifying it is `c(.35, .45, .20)`. Also, when referring to probabilities, the identity link is used:

```{r, tidy = TRUE, eval = FALSE}
formula2 <- c(.35, .45)

dtExp2 <- trtObserve(dtstudy, formulas = formula2, logit.link = FALSE, grpName = "exposure")
```

Figure 2

