---
title: 'simstudy update: adding flexibility to data generation'
author: Package Build
date: '2021-11-09'
slug: []
categories: []
tags:
  - R
  - simstudy
type: ''
subtitle: ''
image: ''
---

```{r, echo=FALSE}
options(digits = 3)
```

A new version of `simstudy` (0.3.0) is now available on [CRAN](https://cran.r-project.org/web/packages/simstudy/index.html){target="_blank"} and on the [package website](https://github.com/kgoldfeld/simstudy/releases){target="_blank"}. Along with some less exciting bug fixes, we have added capabilities to a few existing features: double-dot variable reference, treatment assignment, and categorical data definition. These simple additions should make the data generation process a little smoother and more flexible.
    
## Using non-scalar double-dot variable reference

Double-dot notation was [introduced](https://www.rdatagen.net/post/simstudy-just-got-a-little-more-dynamic-version-0-2-0/) in the last version of `simstudy` to allow data definitions to be more dynamic. Previously, the double-dot variable could only be a scalar value, and with the current version, double-dot notation is now also *array-friendly*. 

Before the examples, here are the necessary packages for this post:

```{r}
library(simstudy)
library(data.table)
library(ggplot2)
```

#### Example 1

In the first example, we want to create a mixture distribution from a vector of values (which we can also do using a *categorical* distribution, more on that in a little bit). We can define the mixture formula in terms of the vector. In this case we are generating permuted block sizes of 2 and 4, specified as

```{r}
sizes <- c(2, 4)
```

The data definition references each element of the vector:

```{r}
defblk <- defData(varname = "blksize", 
   formula = "..sizes[1] | .5 + ..sizes[2] | .5", dist = "mixture")
```

```{r}
genData(1000, defblk)
```

#### Example 2

In this second example, there is a vector variable *tau* of positive real numbers that sum to 1, and we want to calculate the weighted average of three numbers using *tau* as the weights. We could use the following code to estimate a weighted average *theta*:

```{r}
tau <- rgamma(3, 5, 2)
tau <- tau / sum(tau)
tau

d <- defData(varname = "a", formula = 3, variance = 4)
d <- defData(d, varname = "b", formula = 8, variance = 2)
d <- defData(d, varname = "c", formula = 11, variance = 6)
d <- defData(d, varname = "theta", formula = "..tau[1]*a + ..tau[2]*b + ..tau[3]*c", 
  dist = "nonrandom")

set.seed(19483)
genData(4, d)
```

However, we can simplify the calculation of *theta* a bit by using matrix multiplication:

```{r}
d <- updateDef(d, changevar = "theta", newformula = "t(..tau) %*% c(a, b, c)")

set.seed(19483)
genData(4, d)
```

#### Example 3

The arrays can also have **multiple dimensions**, as in a $m \times n$ matrix. If we want to specify the mean outcomes for a $2 \times 2$ factorial study design with two interventions $a$ and $b$, we can use a simple matrix and draw the means directly from the matrix, which in this example is stored in the variable *effect*:

```{r}
effect <- matrix(c(0, 8, 10, 12), nrow = 2)
effect
```

Using double dot notation, it is possible to reference the matrix cell values directly, depending on the values of *a* and *b*:

```{r}
d1 <- defData(varname = "a", formula = ".5;.5", variance = "1;2", dist = "categorical")
d1 <- defData(d1, varname = "b", formula = ".5;.5", 
              variance = "1;2", dist = "categorical")
d1 <- defData(d1, varname = "outcome", formula = "..effect[a, b]",
              variance = 9, dist="normal")
```

```{r}
dx <- genData(1000, d1)
dx
```

The plot shows individual values as well as mean values by intervention arm:

```{r, echo=FALSE}
cbbPalette <- c("#B84226","#B88F26", "#A5B435", "#1B8446",
                "#B87326","#B8A526", "#6CA723", "#1C5974") 

dsum <- dx[, .(outcome=mean(outcome)), keyby = .(a, b)]

ggplot(data = dx, aes(x = factor(a), y = outcome)) +
  geom_jitter(aes(color = factor(b)), width = .2, alpha = .4, size = .2) +
  geom_point(data = dsum, size = 2, aes(color = factor(b))) + 
  geom_line(data = dsum, size = 1, aes(color = factor(b), group = factor(b))) +
  scale_color_manual(values = cbbPalette, name = "  b") +
  theme(panel.grid = element_blank()) +
  xlab ("a")
```

## Assigned treatment using *trtAssign* distribution in `defData`

The function [trtAssign](https://kgoldfeld.github.io/simstudy/articles/treat_and_exposure.html){target="_blank"} currently provides functionality to randomize group assignments using stratification and non-standard ratios (e.g. 2:2:1 for a three arm trial). Starting with version 0.3.0, it is also possible to generate these treatment assignments directly in the `defData` and `genData` process without a separate call to `trtAssign`. We've done this by adding *trtAssign* as a possible distribution. 

In this example, randomization is stratified by *gender* and *age* (specified in the `variance` argument), and randomization is 1:1 treatment to control (specified in `formula`). The outcome *y* is effected by both of these factors as well as the treatment assignment variable *rx*.

```{r}
def <- defData(varname = "male", dist = "binary", 
               formula = .5 , id="cid")
def <- defData(def, varname = "over65", dist = "binary",  
               formula = "-1.7 + .8*male", link="logit")
def <- defData(def, varname = "rx", dist = "trtAssign",
               formula = "1;1", variance = "male;over65")
def <- defData(def, varname = "y", dist = "normal", 
               formula = "20 + 5*male + 10*over65 + 10*rx", variance = 40)

dtstudy <- genData(330, def)
dtstudy
```

Here are the counts and average outcomes for each *gender*, *age*, and *treatment* combination:

```{r}
dtstudy[, .(n = .N, avg = round(mean(y), 1)), keyby = .(male, over65, rx)]
```

## Categogorical data

Finally, in previous versions, the *categorical* distribution generated a set of integer categories:

```{r}
def <- defData(varname = "grp", formula = ".4;.3;.2;.1", dist="categorical")
genData(1000, def)
```

Now, it is possible to generate specific values or string categories by using the `variance` argument:

```{r}
def <- defData(varname = "grp", formula = ".4;.3;.2;.1", 
               variance = "a;b;c;d", dist="categorical")

dd <- genData(1000, def)
dd
```

```{r, echo=FALSE, fig.height = 4}
ggplot(data = dd, aes(x = grp)) +
  geom_bar(width = .5, fill = cbbPalette[5]) +
  theme(panel.grid = element_blank())
```

To replicate Example 1 above, here is an alternative way to generate block sizes of 2 and 4 using the *categorical* distribution and the new functionality.

```{r}
defblk <- defData(varname = "blksize", 
   formula = ".5;.5", variance = "2;4", dist = "categorical")

genData(1000, defblk)
```

## Possible next steps

As we expand the functionality of `simstudy`, we realize that the current structure of the data definition table cannot always easily accommodate all of our new ideas. As a result, we have ended up having to shoehorn some solutions in non-intuitive ways as we grow. We are in the process of reconsidering that structure so that we won't have (as many of) these awkward specifications in the future (though we will be making everything backwards compatible, so no worries there).
