---
title: 'A demo on power estimation by simulation for a cluster randomized trial with a time-to-event outcome'
author: Package Build
date: '2023-05-23'
slug: []
categories: []
tags:
  - R
  - Cluster randomized trials
  - survival analysis
type: ''
subtitle: ''
image: ''
draft: TRUE
---

```{r, echo=FALSE}
options(digits = 2)
```

```{r, message=FALSE}
library(simstudy)
library(data.table)
library(survival)
library(survminer)
library(coxme)
library(parallel)
```

### Defining shape of survival curve

```{r}
points <- list(c(30, 0.90), c(365, .50))
r <- survGetParams(points)
r
```

```{r survparam, fig.height=3.5, fig.width = 5.5}
survParamPlot(r[1], r[2], points = points, n = 1000, limits = c(0, 365) )
```

### Evaluating cluster variation 

#### Visualizing cluster variation

```{r}
defc <- defData(varname = "b", formula = 0, variance = 0.1)

defa <- defDataAdd(varname = "start_day", formula = "1;182", dist = "uniformInt")
defa <- defDataAdd(defa, varname = "censor", 
  formula = "365 - start_day ", dist = "nonrandom")

defs <- defSurv(varname = "ttc", formula = "r[1] + b", shape = r[2])
```

```{r}
set.seed(589823)

dc <- genData(20, defc, id = "site")
dd <- genCluster(dc, "site", numIndsVar = 200, "id")
dd <- addColumns(defa, dd)
dd <- genSurv(dd, defs, digits = 0)
dd <- addCompRisk(dd, events = c("ttc", "censor"), 
  timeName = "time", censorName = "censor", keepEvents = TRUE)
```

```{r, fig.height=3.5, fig.width = 5.5}
fit <- survfit(Surv(time, event) ~ site, data = dd)

j1 <- ggsurvplot(fit, data = dd, title = "Cluster variance = 0.100")
j1$plot <- j1$plot + theme(legend.position = "none",
                           plot.title = element_text(size = 11, face = "bold"))
```


```{r variation, echo = FALSE, fig.height=4, fig.width = 10}
library(ggpubr)

defc <- defData(varname = "b", formula = 0, variance = 0.005)

dc <- genData(20, defc, id = "site")
dd <- genCluster(dc, "site", numIndsVar = 200, "id")
dd <- addColumns(defa, dd)
dd <- genSurv(dd, defs, digits = 0)
dd <- addCompRisk(dd, events = c("ttc", "censor"), 
  timeName = "time", censorName = "censor", keepEvents = TRUE)

fit <- survfit(Surv(time, event) ~ site, data = dd)

j2 <- ggsurvplot(fit, data = dd, title = "Cluster variance = 0.005")

j2$plot <- j2$plot + theme(legend.position = "none",
                           plot.title = element_text(size = 11, face = "bold"))

ggarrange(j1$plot, j2$plot, ncol = 2, nrow = 1)
```

#### Variation of the probability of an event across clusters

```{r histprob, warning=FALSE, fig.width=4, fig.height=3}
defc <- defData(varname = "b", formula = 0, variance = 0.1)

defa <- defDataAdd(varname = "start_day", formula = "1;182", dist = "uniformInt")
defa <- defDataAdd(defa, varname = "censor", 
  formula = "365 - start_day ", dist = "nonrandom")

defs <- defSurv(varname = "ttc", formula = "r[1] + b", shape = r[2])

dc <- genData(1000, defc, id = "site")
dd <- genCluster(dc, "site", numIndsVar = 500, "id")
dd <- addColumns(defa, dd)
dd <- genSurv(dd, defs, digits = 0)

dd[, event := 1 * (ttc <= 365)]

ds <- dd[, .(p = mean(event)), keyby = site]

ggplot(data = ds, aes(x = p)) +
  geom_histogram(binwidth = .05) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 200)) +
  ggtitle("Cluster variance = 0.100") +
  xlab("observed proportion of events") +
  theme(panel.grid = element_blank(),
        plot.title = element_text(size = 11, face = "bold")) 
```


```{r varprob, echo=FALSE, warning=FALSE, message=FALSE, fig.width=6, fig.height=4}
s_generate <- function(nsite, ninds, s2) {
  
  defc <- defData(varname = "b", formula = 0, variance = s2)

  dc <- genData(nsite, defc, id = "site")
  dd <- genCluster(dc, "site", numIndsVar = ninds, "id")
  dd <- addColumns(defa, dd)
  dd <- genSurv(dd, defs, digits = 0)

  dd[, event := 1 * (ttc <= 365)]

  dd[, .(p = mean(event)), keyby = site]
}

replicate <- function(s2) {
  ds <- s_generate(1000, 500, s2)
  data.table(s2 = s2, min_p = quantile(ds$p, prob = 0.025), 
             med_p = quantile(ds$p, prob = 0.5),
             max_p = quantile(ds$p, prob = 0.975))
}

res <- parallel::mclapply(seq(0.005, 0.1, by = .005), function(a) replicate(a))
dp <- rbindlist(res)

ggplot(data = dp, aes(x = s2, y = med_p)) +
  geom_point(size = .8) +
  geom_errorbar(aes(ymin = min_p, ymax = max_p), width = .0, color = "grey65") +
  theme(panel.grid = element_blank()) +
  scale_x_continuous(
    name = "\nvariance of random effect in survival data generation",
    breaks = seq(0, .1, by = .01)
  ) +
  scale_y_continuous(
    name = "probability of event",
    limits = c(0.25, 0.75)
  )
```

### Evaluating the effect size

```{r}
defa <- defData(varname = "rx", formula = "1;1", dist = "trtAssign")
defa <- defData(defa, varname = "start_day", formula = "1;182", dist = "uniformInt")
defa <- defDataAdd(defa, varname = "censor", 
  formula = "365 - start_day ", dist = "nonrandom")

defs <- defSurv(varname = "ttc", formula = "r[1] + 0.4 * rx", shape = r[2])

dd <- genData(600, defa)
dd <- genSurv(dd, defs, digits = 0)
dd <- addCompRisk(dd, events = c("ttc", "censor"), 
  timeName = "time", censorName = "censor", keepEvents = TRUE)
```


```{r effect, fig.height=3.5, fig.width=5.5}
fit <- survfit(Surv(time, event) ~ rx, data = dd)

ggsurvplot(
  fit, 
  data = dd, 
  legend.title = "", 
  legend = c(.25, 0.25),
  palette = "jco"
)
```


```{r}
dd[, event := 1 * (ttc <= 365)]
dd[, .(p = mean(event)), keyby = rx]
```

This is an odds ratio of 1.80, risk ratio of 1.26, and risk difference of 14 percentage points.

### Complete data generation and model estimation

```{r}
defc <- defData(varname = "b", formula = 0, variance = 0.05)
defc <- defData(defc, varname = "rx", formula = "1;1", dist = "trtAssign")

defa <- defDataAdd(varname = "start_day", formula = "1;182", dist = "uniformInt")
defa <- defDataAdd(defa, varname = "censor", 
  formula = "365 - start_day ", dist = "nonrandom")

defs <- defSurv(varname = "ttc", formula = "r[1] + 0.4 * rx + b", shape = r[2])
```

```{r, eval=FALSE}
dc <- genData(1000, defc, id = "site")
dd <- genCluster(dc, "site", numIndsVar = 500, "id")
dd <- addColumns(defa, dd)
dd <- genSurv(dd, defs, digits = 0)
dd <- addCompRisk(dd, events = c("ttc", "censor"), 
  timeName = "time", censorName = "censor", keepEvents = TRUE)

fit_coxme <-coxme(Surv(time, event) ~ rx + (1 | site), data = dd)

summary(fit_coxme)
```

```{r, echo=FALSE}
load("data/coxfit.rda")
fit_coxme
```

### Power estimation

Check out the <a href="#addendum">addendum</a> for code details.

```{r, echo = FALSE}
extract_coxme_table <- function (mod) {
  beta <- mod$coefficients 
  nvar <- length(beta)
  nfrail <- nrow(mod$var) - nvar
  se <- sqrt(diag(mod$var)[nfrail + 1:nvar])
  z <- round(beta/se, 2)
  p <- signif(1 - pchisq((beta/se)^2, 1), 2)
  table=data.table(beta = beta, se = se, z = z, p = p)
  return(table)
}

s_def <- function() {
  
  defc <- defData(varname = "b", formula = 0, variance = "..s2")
  defc <- defData(defc, varname = "rx", formula = "1;1", dist = "trtAssign")
  
  defa <- defDataAdd(varname = "start_day", formula = "1;182", dist = "uniformInt")
  defa <- defDataAdd(defa, varname = "censor", 
                     formula = "365 - start_day ", dist = "nonrandom")
  
  defs <- defSurv(varname = "ttc", formula = "-4.815 + 0.4 * rx + b", shape = 1.326)
  
  defa2 <- defDataAdd(varname = "event6", 
                      formula = "1*(ttc <= 182)", dist = "nonrandom")
  
  return(list(defc = defc, defa = defa, defs = defs, defa2 = defa2))
  
}

s_generate <- function(argsvec, list_of_defs) {
  
  list2env(list_of_defs, envir = environment())
  list2env(as.list(argsvec), envir = environment())
  
  dc <- genData(nsites, defc, id = "site")
  dd <- genCluster(dc, "site", ninds, "id")
  dd <- addColumns(defa, dd)
  dd <- genSurv(dd, defs, digits = 0)
  dx <- addCompRisk(dd, events = c("ttc", "censor"), 
                    timeName = "time", censorName = "censor", keepEvents = TRUE)
  dx <- addColumns(defa2, dx)
  
  dx[]
  
}

s_replicate <- function(argsvec, list_of_defs) {
  
  dx <- s_generate(argsvec, list_of_defs)
  
  coxfitm <-coxme(Surv(time, event) ~ rx + (1 | site), data = dx)
  
  list2env(as.list(argsvec), envir = environment())
  
  return(data.table(
    nsites = nsites,
    ninds = ninds,
    s2 = s2,
    est_s = fixef(coxfitm), 
    re.var_s = VarCorr(coxfitm)$site,
    p_s = extract_coxme_table(coxfitm)$p
  ))
  
}

s_scenarios <- function(argsvec, nreps) {
  
  list_of_defs <- s_def()
  
  rbindlist(
    parallel::mclapply(
      X = 1 : nreps, 
      FUN = function(x) s_replicate(argsvec, list_of_defs), 
      mc.cores = 4)
  )
  
}
```

```{r}
scenario_list <- function(...) {
  argmat <- expand.grid(...)
  return(asplit(argmat, MARGIN = 1))
}

nsites <- c(20, 30)
ninds <- c(15)
s2 <- c(0.03, 0.04)

scenarios <- scenario_list(nsites = nsites, ninds = ninds, s2 = s2)
model.ests <- mclapply(scenarios, function(a) s_scenarios(a, nrep = 3))

model.ests
```

```{r power, echo = FALSE, fig.height = 2.75, fig.width = 8}
load("data/ss.rda")

resum[, `number of sites` := nsites]

ggplot(data = resum, mapping = aes(x = s2, y = p_m)) +
  geom_hline(yintercept = 0.8, color = "white", linewidth = 1.2) +
  geom_line(aes(color = factor(ninds), group = ninds), linewidth = 0.9) +
  facet_grid(. ~ `number of sites`, labeller = label_both) +
  theme(
    panel.grid = element_blank(),
    axis.text = element_text(size = 8),
    panel.spacing = unit(0.65, "lines")
  ) +
  scale_color_brewer(
    name = "size", 
    guide = guide_legend(reverse = TRUE), 
    type = "div", 
    palette = "Dark2"
  ) +
  ylab("estimated power") +
  xlab("\nvariance of random effect")

```

<a name="addendum"></a>  

\ 

## Addendum

```{r, eval = FALSE}
extract_coxme_table <- function (mod) {
  beta <- mod$coefficients 
  nvar <- length(beta)
  nfrail <- nrow(mod$var) - nvar
  se <- sqrt(diag(mod$var)[nfrail + 1:nvar])
  z <- round(beta/se, 2)
  p <- signif(1 - pchisq((beta/se)^2, 1), 2)
  table=data.table(beta = beta, se = se, z = z, p = p)
  return(table)
}

s_def <- function() {
  
  defc <- defData(varname = "b", formula = 0, variance = "..s2")
  defc <- defData(defc, varname = "rx", formula = "1;1", dist = "trtAssign")
  
  defa <- defDataAdd(varname = "start_day", formula = "1;182", dist = "uniformInt")
  defa <- defDataAdd(defa, varname = "censor", 
                     formula = "365 - start_day ", dist = "nonrandom")
  
  defs <- defSurv(varname = "ttc", formula = "-4.815 + 0.4 * rx + b", shape = 1.326)
  
  defa2 <- defDataAdd(varname = "event6", 
                      formula = "1*(ttc <= 182)", dist = "nonrandom")
  
  return(list(defc = defc, defa = defa, defs = defs, defa2 = defa2))
  
}

s_generate <- function(argsvec, list_of_defs) {
  
  list2env(list_of_defs, envir = environment())
  list2env(as.list(argsvec), envir = environment())
  
  dc <- genData(nsites, defc, id = "site")
  dd <- genCluster(dc, "site", ninds, "id")
  dd <- addColumns(defa, dd)
  dd <- genSurv(dd, defs, digits = 0)
  dx <- addCompRisk(dd, events = c("ttc", "censor"), 
                    timeName = "time", censorName = "censor", keepEvents = TRUE)
  dx <- addColumns(defa2, dx)
  
  dx[]
  
}

s_replicate <- function(argsvec, list_of_defs) {
  
  dx <- s_generate(argsvec, list_of_defs)
  
  coxfitm <-coxme(Surv(time, event) ~ rx + (1 | site), data = dx)
  
  list2env(as.list(argsvec), envir = environment())
  
  return(data.table(
    nsites = nsites,
    ninds = ninds,
    s2 = s2,
    est_s = fixef(coxfitm), 
    re.var_s = VarCorr(coxfitm)$site,
    p_s = extract_coxme_table(coxfitm)$p
  ))
  
}

s_scenarios <- function(argsvec, nreps) {
  
  list_of_defs <- s_def()
  
  rbindlist(
    parallel::mclapply(
      X = 1 : nreps, 
      FUN = function(x) s_replicate(argsvec, list_of_defs), 
      mc.cores = 4)
  )
  
}
```

