---
title: Flexible simulation in simstudy with customized distribution
  functions
author: Package Build
date: '2022-08-30'
slug: []
categories: []
tags:
  - R
  - simulation
type: ''
subtitle: ''
image: ''
draft: TRUE
---

Really, the only problem with the `simstudy` package (&#128516;) is that there is a limit to the possible probability distributions that are available (the current count is 15 - see [here](https://kgoldfeld.github.io/simstudy/articles/simstudy.html){target="_blank"} for a complete description). Actually, it turns out that there is more flexibility than first meets the eye, and we can easily accommodate a limitless number as long as you are willing to provide some extra code.

I am going to illustrate this with two examples, the first to implement a truncated normal distribution, and the second the implement the flexible non-linear data generating algorithm that I [described last time](https://www.rdatagen.net/post/2022-08-09-simulating-data-from-a-non-linear-function-by-specifying-some-points-on-the-curve/){target="_blank"}. 

Before we get going, here are the necessary libraries:

```{r, message=FALSE, warning-FALSE}
library(simstudy)
library(data.table)
library(msm)
library(ggplot2)
library(mgcv)
```

#### General concept

In the data definition step, it is possible to specify any valid `R` function in the *formula* argument. If *dist* is specified as "nonrandom", then `simstudy` will generate data based on that function. (Yes, the specification as "nonrandom" is a bit awkward since we are defining a stochastic data generating process in this case; in future versions I plan to allow *dist* to be specified as "custom" to make this less dissonant.)

In this example, I want to be able to generate data from a truncated normal distribution. There is an existing function `rtnorm` in the `msm` package that I can take advantage here. What I have done is essentially create a wrapper function that makes a single draw from the truncated distribution with a specified mean, standard deviation, and pair of truncation bounds:

```{r}
trunc_norm <- function(mean, sd, lower, upper) {
  rtnorm(n = 1, mean = mean, sd = sd, lower = lower, upper = upper)
}
```

Now that `trunc_norm` has been created, I am free to use this is in a data definition statement. And even more important, the call to `trunc_norm` can depend on other variables; in this case, I have created binary variable *x* that will determine the upper and lower bounds of the distribution. When $x=0$, the $N(0, 3.5^2)$ distribution is truncated at -5 and 5, and when $x=1$, the distribution is truncated at -8 and 8.

```{r}
defI <- defData(varname = "x", formula = 0.5, dist = "binary")
defI <- defData(defI, varname = "y", 
  formula = "trunc_norm(mean = 0, sd = 3.5, 
               lower = -5 + -3*x, upper = 5 + 3*x)",
  dist = "nonrandom")
```

The generated data appear to have the properties that we would expect:

```{r}
dd <- genData(1000, defI)
```

```{r, echo=FALSE, fig.width = 5, fig.height=3}
ggplot(data = dd, aes(x = factor(x), y = y)) +
  geom_jitter(size = .4, color = "#9f390c", width = .3) +
  scale_y_continuous(limits = c(-9, 9), breaks = c(-8, -5, 0, 5, 8)) +
  xlab("x") +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())
```

#### Application to non-linear data generation

[Last time](https://www.rdatagen.net/post/2022-08-09-simulating-data-from-a-non-linear-function-by-specifying-some-points-on-the-curve/){target="_blank"}, I described an approach to generate a variable that has a non-linear response with another variable. At the end of that post, I created a couple of functions, one of which can be referred to in the `defData` statement to generate the data. (I plan on implementing these functions in `simstudy`, but I was eager to get the concept out there in case any one has some suggestions.)

```{r, echo = FALSE}

interp_line <- function(x, y, xend, yend, npoints = 100) {
  slope <- (yend - y)/(xend - x)
  b <- y - slope*x
  newx <- seq(x, xend, length = npoints)
  newy <- newx*slope + b
  
  data.table::data.table(x = newx, y = newy)
}

getNLfunction <- function(dd, k = 0, show.plot = TRUE) {
  dc <- copy(dd)
  
  dc[, xend := (shift(x, type = "lead"))]
  dc[, yend := (shift(y, type = "lead"))]
  
  dc <- dc[ -.N]
  dc[, id := .I]
  
  dx <- dc[, interp_line(x, y, xend, yend, 100), keyby = id]
  
  gam.fit <- gam(y ~ s(x, k = nrow(dd) + k), data = dx)
  dx[, ypred := predict(gam.fit)]
  
  if (show.plot) {
    
    p <- ggplot(data = dx, aes(x=x, y = ypred)) +
      geom_line()  +
      geom_point(data = dd, aes(x = x, y = y)) +
      theme(panel.grid = element_blank()) +
      ylab("y")
    
    print(p)
  }
  
  return(gam.fit)

}

genNL <- function(nf, x, sd) {
  predict(nf, newdata= data.frame(x = x))  + rnorm(length(x), 0, sd)
}
```

In the first step, I need to generate a smooth function by specifying a few points. I do this by calling `getNLfunction`. (If you want the code for this, let me know, but I actually provided most of it last week.) The variable `nlf` is an object that contains the function:

```{r, fig.height=3, fig.width=5}
dpoints <- data.table(x = c(20, 30, 53, 65, 80), y = c(15, 44, 60, 55, 35))
nlf <- getNLfunction(dpoints)
```

The function `genNL` makes predictions based on the `nlf` object and adds a little Gaussian noise. We use the same approach as we did above for the truncated normal to generate different responses $y$ based on the level of $x$:

```{r, fig.height = 3, fig.width = 5}
def <- defData(varname = "x", formula = "20;80", dist = "uniform")
def <- defData(def, varname = "y", 
  formula = "genNL(nf = ..nlf, x, sd = 10)", dist = "nonrandom")

dd <- genData(300, def)
```

```{r, fig.height = 3, fig.width = 5, echo=FALSE}
ggplot(data = dd, aes(x=x, y = y)) +
  geom_point(size = .4, color = "#9f390c")  +
  theme(panel.grid = element_blank())
```

And if we introduce much less noise, we get much closer to the original underlying function specified by our points:

```{r, fig.height = 3, fig.width = 5}
def <- defData(varname = "x", formula = "20;80", dist = "uniform")
def <- defData(def, varname = "y", 
  formula = "genNL(nf = ..nlf, x, sd = 0.5)", dist = "nonrandom")

dd <- genData(300, def)
```

```{r, fig.height = 3, fig.width = 5, echo=FALSE}
ggplot(data = dd, aes(x=x, y = y)) +
  geom_point(size = .4, color = "#9f390c")  +
  theme(panel.grid = element_blank())
```
